A generic and fast C++ optimization framework

### 2017-10-21 paper.tex
Refactor paper to make it 7 pages, finish unfinished sections.
The basic tricks I used to get the size down:

 * Add \vspace*{-0.5em} (or whatever value) to remove vertical space---but not
   so much the paper is cramped.
 * Inline code in a few places to avoid extra line breaks.
 * Use \footnotesize for code listings, which hopefully aren't too small.
 * Put code directly into paper instead of into a listing that LaTeX can't split
   across pages.
 * Find different words or shorter phrases to remove the last line of paragraphs
   where the line was short.
 * Tilt column headers in Table 1 to

% to compile a camera-ready version, add the [final] option, e.g.:
% \usepackage[final]{nips_2017}

@@

\newcolumntype{R}[2]{%
  >{\adjustbox{angle=#1,lap=\width-(#2)}\bgroup}%
  l%
  <{\egroup}%
}
\newcommand*\rot{\multicolumn{1}{R{60}{2em}}}

@@

\author{
  Shikhar Bhardwaj \\
  \AND
  Yannis Mentekidis \\
  \AND
  Marcus Edel \\
  Free University of Berlin \\
  Arnimallee 7, 14195 Berlin \\
  \texttt{marcus.edel@fu-berlin.de} \\
  \AND
  Ryan R. Curtin \\
  Symantec Corporation \\
  \texttt{ryan@ratml.org} \\
}

@@

\begin{abstract}
The development of the {\bf mlpack} C++ machine learning library
(\url{http://www.mlpack.org/}) has required the design and implementation of a
flexible, robust optimization system that is able to solve the types of
arbitrary optimization problems that may arise all throughout machine learning
problems.  You care because:

\begin{itemize}
  \item It's easy to implement new optimizers.
  \item It's easy to implement new functions to be optimized.
  \item This system is more comprehensive than other techniques that may focus
only on neural networks or single problems.
  \item This system is also more specific and offers more room for optimization
than super general systems like scipy.optimize().
\end{itemize}

This abstract will need to be rewritten.
\end{abstract}

@@

However, the current landscape of optimization frameworks for machine learning
is not particularly comprehensive.  A variety of tools such as
Caffe \cite{jia2014caffe},
% mxnet ??
TensorFlow \cite{abadi2016tensorflow},
and
Keras \cite{chollet2015}
have optimization frameworks, but they are limited to SGD-type optimizers and
are only able to optimize deep neural networks or related structures.  Thus
expressing arbitrary machine learning objective functions can be difficult or in
some cases not possible.  Other libraries, like
% Shogun \cite{shogun}
%and
scikit-learn \cite{pedregosa2011scikit},
do have optimizers implementation, but generally not in a coherent framework and
often the implementations may be specific to an individual machine learning
algorithm.  At a higher level, many programming languages may have generic
optimizers, like SciPy \cite{jones2014scipy} and MATLAB, but typically these
optimizers are not suitable for large-scale machine learning tasks where, e.g.,
calculating the full gradient of all of the data may not be feasible.  Table
\ref{tab:features} provides a rough overview of the capabilities of many common
machine learning libraries and languages.

Given this situation, we have developed a flexible optimization infrastructure
in the {\bf mlpack} C++ machine learning library \cite{mlpack2013}.  This
infrastructure makes it easy to combine nearly any type of optimizer with nearly
any type of objective function, and has allowed us to minimize the effort
necessary to both implement new optimizers and to implement new machine learning
algorithms that depend on optimization.  Since the framework is implemented in
C++ and uses template metaprogramming, we are able to preserve a clean syntax
while simultaneously allowing compile-time optimizations that can result in a
generic optimizer that is equivalently fast to an optimizer written specifically
for one objective function.

@@

\vspace*{-1.5em}

@@

\section{The mlpack machine learning library}

@@

\begin{minted}[fontsize=\footnotesize]{c++}
  template<bool UseKernel, typename KernelType, typename MatType>
  class MeanShift;
\end{minted}

Thus a user wishing to perform kernelized mean shift clustering with the
Gaussian kernel might simply use the class

\begin{minted}[fontsize=\footnotesize]{c++}
  MeanShift<true, GaussianKernel, arma::mat>
\end{minted}

\noindent where {\tt arma::mat} is Armadillo's dense matrix type and {\tt
GaussianKernel} is a class that provides one simple {\tt Evaluate()} function
that returns the kernel value between its two arguments.  Since the {\tt
KernelType} argument is a template argument, any class can be used---it does not
need to be functionality supplied with mlpack.  Therefore, a user can simply
implement their own {\tt KernelType} class with the single {\tt Evaluate()}
function and use that as a template argument to the {\tt MeanShift} class, and
the compiler will generate a specialized implementation using the custom kernel.
Similar support could be accomplished by, e.g., function pointers (and this is
often done in other languages); however, templates allow us to avoid the runtime
cost of dereferencing the function pointer by generating code that directly
calls the correct function. This cost can be non-negligible, especially in
situations where the function is called many times, such as the {\tt Evaluate()}
function of an optimization problem.

@@

\section{Requirements for optimizers and functions}

@@

\begin{equation}
\operatorname{argmin}_{x} f(x)
\end{equation}

\noindent for any function $f(x)$ that takes some vector input $x$.  However, it
is impossible to implement something both so generic and fast---for instance,
gradient-based approaches converge far more quickly than gradient-free
approaches (in general), so we must design an abstraction that is able to
simultaneously generalize to many problem types, as well as able to take
advantage of accelerations and optimizations.

Let us split the class of functions to be optimized into some subclasses:

\begin{itemize}
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  \item {\bf numeric}: all elements of $x$ take values in $\mathcal{R}$
  \item {\bf sparse}: the gradient $f'(x)$ or $f_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the gradient $f'_j(x)$ is computable for
individual elements $x_j$ of $x$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}

@@

In some cases, some subclasses of functions can still be optimized with more
general optimizers.  For instance, separable objective functions can still be
optimized with an optimizer that does not specifically support separable
functions.  Similarly, a sparse differentiable function may be optimized with
any optimizer that can handle differentiable functions; however, that optimizer
may not take advantage of the sparseness.

Now, with an understanding of the types of functions that we wish to support and
their individual characteristics, we can develop an API that a given function
should implement the relevant parts of.

\section{{\tt FunctionType} API}

In order to facilitate consistent implementations, we have defined a {\tt
FunctionType} API that describes all the functions that an objective function
may (or must) implement, depending on the subclass of objective function.

At the time of writing, we offer a few implementations of this API, each
designed to cover a special case.

\begin{itemize}
  \item{\tt FunctionType}: Requires the implementation of {\tt Evaluate()} and
  {\tt Gradient()}. This is a generic function that is used by the
  {\tt GradientDescent} optimizer.
  \item{\tt DecomposableFunctionType}: In addition to {\tt FunctionType}'s
  requirements, requires the implementation of {NumFunctions()}, which returns
  the number of parts the optimization problem can be decomposed into. The other
  two functions ({\tt Gradient(), Evaluate()}) accept an index as argument and
  only evaluate one of the decomposable parts. This is used by our {\tt AdaGrad}
  implementations, some {\tt SGD} implementations, {\tt RMSProp} and
  {\tt SMORMS3}.
  \item{\tt SparseFunctionType}: Similar to {\tt DecomposableFunctionType},
  however this function type returns a sparse gradient, which makes it ideal for
  sparse, parallelizable optimization problems. This function type is used by
  {\bf mlpack}'s parallel implementation of SGD, {\tt ParallelSGD}.
  \item{\tt LagrangianFunctionType}: This function type fits constrained
  optimization problems. It requires the implementation of {\tt NumConstraints},
  {\tt EvaluateConstraint()} and {\tt GradientConstraint}, which are required by
  optimizers like {\tt AugLagrangian} and {\tt LRSDP}.
\end{itemize}

In addition to these function types, certain optimizers require specific
functionality to achieve their goal. The {\tt FunctionType} API offers users the
flexibility to define a custom function signature. As an example, {\bf mlpack}'s
implementation of L-BFGS requires a function that exposes a
{\tt GetInitialPoint()} function.

{\bf <TODO: we can add a list of each function and what it does.  A basic list
is below.  We can elaborate on this significantly.>}

\begin{itemize}
  \item {\tt Evaluate()}: Evaluate a function with the given function parameter.

  \item {\tt Gradient()}: Evaluate the gradient of a function with the given
      function parameter and return the gradient.

  \item {\tt PartialGradient()}: Evaluate the partial (sparse) gradient of a
      function with the given parameter and return the partial gradient.

  \item ...
  \item constraints?
  \item discrete variable support?
  \item batch support
\end{itemize}

A user may implement as many (or as few) of these functions as they would like,
and this will allow them to use a larger set of optimizers.

@@

% \vspace*{-0.3em}
\begin{listing}[H]
\begin{minted}[fontsize=\footnotesize]{c++}
class ObjectiveFunction {
 ObjectiveFunction() {
  in = arma::vec("20 12 15 100");
  bi = arma::vec("-4 -2 -3 -8");
 }

 void Shuffel() { order = arma::shuffle(arma::linspace<arma::uvec>(0, 3, 4)); }

 size_t NumFunctions() { return 4; }

 double Evaluate(const arma::mat& para, size_t s, size_t bs) {
  double cost = 0;
  for (size_t i = s; i < s + bs; i++)
   cost += std::pow(para(i), 2) + bi(order(i)) * para(order(i)) + in(order(i));
  return cost;
 }

 void Gradient(const arma::mat& para, size_t s, size_t bs, arma::mat& g) {
  g.zeros(arma::size(p));
  for (size_t i = s; i < s + bs; i++)
   g(i) += 2 * para(i) + bi(order(i));
  g /= bs;
 }

 arma::vec in, bi; arma::uvec order;
}
\end{minted}
\vspace{-0.6em}
\caption{Simple objective function where each dimension has a parabola with a
  distinct minimum}
\vspace{-0.8em}
\end{listing}
\vspace{-0.8em}
\begin{listing}[H]
\begin{minted}[fontsize=\footnotesize]{c++}
class SimpleOptimizer {
 SimpleOptimizer(size_t bs = 2, double lr = 0.1) : bs(bs), lr(lr) { }

 template<typename FunctionType>
 double Optimize(FunctionType& function, arma::mat& parameter) {
  arma::mat gradient;
  for (size_t i = 0; i < 300; ++i) {
   const size_t funIdx = i % 4;
   function.Gradient(parameter, funIdx, gradient, batchSize);
   parameter -= lr * gradient;
   function.Shuffel();
  }
  return function.Evaluate(parameter, 0, function.NumFunctions());
 }
 size_t bs;  double lr;
}
\end{minted}
\vspace{-0.6em}
\caption{Simple stochastic gradient descent method with a static number of
iterations to optimize the function parameter ({\tt bs} defines the batch size
and {\tt lr} the learning rate). )}
\vspace{-0.6em}
\end{listing}
\vspace{-0.8em}
Using this code, we can compile the code, and when we run it

\begin{minted}[fontsize=\footnotesize]{c++}
ObjectiveFunction function; arma::mat parmeter("0 0 0 0;")
SimpleOptimizer optimizer;
std::cout << optimizer.Optimize(function, parameter);
\end{minted}
we receive the following output:

xxxx

@@

\section{Static checking of functions}

{\bf <TODO: probably a better section title>}

Unfortunately, template metaprogramming can result in some very lengthy error
messages.  Therefore, we must be careful to ensure that a user is able to easily
debug a problem when they implement an objective function or gradient.  To
provide good error message output, we can use C++'s template metaprogramming
support to determine what methods a type has available.  Similarly, we can also
use static compile-time constants to denote the methods that are require by a
specific optimizer.  This is implemented via SFINAE \cite{sfinae}.  A {\tt
static\_assert()} is raised when a given objective function does not have the
required methods for a given optimizer.  {\bf <TODO: implement this, so we are
not lying.  But it should be easy enough, maybe an hour of work max.>}

For instance, when attempting to use the L-BFGS optimizer without having a
{\tt Gradient()} function implemented, the user will receive the following error
message:

{\bf <TODO: put error message here.  If we are short on space, let's cut this
example and make this section shorter.>}

@@

% I guess to save some space we should group them.
\begin{itemize}
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

  \item {\bf Other:} Conditional Gradient Descent (Frank-Wolfe algorithm)

  \item {\bf Objective functions:} Neural Networks, Logistic regression,
      Matrix completion, Neighborhood Components Analysis, Regularized SVD,
      Reinforcement learning, Softmax regression, Sparse autoencoders,
      Sparse SVM
\end{itemize}

@@

We have identified that the support for generic and robust optimization is not
currently available in most machine learning toolkits, and acted upon this
observation to provide an easy framework for both implementing new optimizers
and new objective functions to be optimized inside of the {\bf mlpack} machine
learning library.

The framework provided by {\bf mlpack} supports a wide array of special cases,
and offers specialized algorithms that outperform their classic generic
 (alternatives.

>>>>

% to compile a camera-ready version, add the [final] option, e.g.:
%\usepackage[final]{nips_2017}

@@

\newcolumntype{R}[2]{%
  >{\adjustbox{angle=#1,lap=\width-(#2)}\bgroup}%
  l%
  <{\egroup}%
}
\newcommand*\rot{\multicolumn{1}{R{30}{2em}}}

@@

\author{
  Shikhar Bhardwaj \\
  \AND
  Yannis Mentekidis \\
  \AND
  Marcus Edel \\
  Free University of Berlin \\
  Arnimallee 7, 14195 Berlin \\
  \texttt{marcus.edel@fu-berlin.de} \\
  \AND
  Ryan R. Curtin \\
  Symantec Corporation \\
  Atlanta, GA, USA 30318 \\
  \texttt{ryan@ratml.org} \\
}

@@

\begin{abstract}
The development of the {\bf mlpack} C++ machine learning library
(\url{http://www.mlpack.org/}) has required the design and implementation of a
flexible, robust optimization system that is able to solve the types of
arbitrary optimization problems that may arise all throughout machine learning
problems.  In this paper we present the generic optimization framework that we
have designed for mlpack.  A key priority in the design was ease of
implementation of both new optimizers and new objective functions to be
optimized; therefore, implementation of a new optimizer requires only one
method and implementation of a new objective function requires at most four
functions.  This leads to simple and intuitive code, and for fast prototyping
and experimentation, this is of paramount importance.  When compared to
optimization frameworks of other libraries, we find that mlpack's supports more
types of objective functions, is able to make optimizations that other
frameworks do not, and seamlessly supports user-defined objective functions
{\it and} optimizers.
%\begin{itemize}
%  \item It's easy to implement new optimizers.
%  \item It's easy to implement new functions to be optimized.
%  \item This system is more comprehensive than other techniques that may focus
%only on neural networks or single problems.
%  \item This system is also more specific and offers more room for optimization
%than super general systems like scipy.optimize().
%\end{itemize}
%
%This abstract will need to be rewritten.
\end{abstract}

@@

However, the current landscape of optimization frameworks for machine learning
is not particularly comprehensive.  A variety of tools such as
Caffe \cite{jia2014caffe},
% mxnet ??
TensorFlow \cite{abadi2016tensorflow},
and
Keras \cite{chollet2015}
have optimization frameworks, but they are limited to SGD-type optimizers and
are only able to optimize deep neural networks or related structures.  Thus
expressing arbitrary machine learning objective functions can be difficult or in
some cases not possible.  Other libraries, like
% Shogun \cite{shogun}
%and
scikit-learn \cite{pedregosa2011scikit},
do have optimizers, but generally not in a coherent framework and
often the implementations may be specific to an individual machine learning
algorithm.  At a higher level, many programming languages may have generic
optimizers, like SciPy \cite{jones2014scipy} and MATLAB, but typically these
optimizers are not suitable for large-scale machine learning tasks where, e.g.,
calculating the full gradient of all of the data may not be feasible.  Table
\ref{tab:features} provides a rough overview.

Given this situation, we have developed a flexible optimization infrastructure
in the {\bf mlpack} C++ machine learning library \cite{mlpack2013}.  This
infrastructure makes it easy to combine nearly any type of optimizer with nearly
any type of objective function, and has allowed us to minimize the effort
necessary to both implement new optimizers and to implement new machine learning
algorithms that depend on optimization.  Since the framework is implemented in
C++ and uses template metaprogramming, we are able to preserve clean syntax
while simultaneously allowing compile-time optimizations that can result in a
generic optimizer that is as fast as an optimizer written for one specific objective function.

@@

\vspace*{-1.8em}

@@

\vspace*{-0.3em}
\section{The mlpack machine learning library}
\vspace*{-0.2em}

@@

\vspace*{-0.4em}
\begin{minted}[fontsize=\footnotesize]{c++}
  template<bool UseKernel, typename KernelType, typename MatType>
  class MeanShift;
\end{minted}
\vspace*{-0.4em}

Thus a user wishing to perform kernelized mean shift clustering with the
Gaussian kernel might simply use the class
{\footnotesize \mintinline{c++}{MeanShift<true, GaussianKernel, arma::mat>}}
\noindent where {\tt arma::mat} is Armadillo's dense matrix type and {\tt
GaussianKernel} is a class that provides one simple {\tt Evaluate()} method.

Since the {\tt KernelType} argument is a template argument, any class can be
used---it does not need to be part of mlpack.  Therefore, a user can simply
implement their own {\tt KernelType} class with the single {\tt Evaluate()}
method and use that as a template argument to the {\tt MeanShift} class, and
the compiler will generate a specialized implementation using the custom
kernel.  Similar support could be accomplished by, e.g., function pointers (and
this is often done in other languages); however, templates allow us to avoid
the runtime cost of dereferencing the function pointer by generating code that
directly calls the correct method. This cost can be non-negligible, especially
in situations where the method is called many times, such as the {\tt
Evaluate()} method of an optimization problem.

@@

\vspace*{-0.3em}
\section{Requirements for optimizers and functions}
\vspace*{-0.2em}

@@

\vspace*{-0.5em}
\begin{equation}
\operatorname{argmin}_{x} f(x)
\end{equation}
\vspace*{-1.3em}

\noindent for any function $f(x)$ that takes some vector input $x$.  But it
is impossible to implement something both so generic and fast---for instance,
gradient-based approaches converge far more quickly than gradient-free
approaches (in general), so we must design an abstraction that is able to
simultaneously generalize to many problem types, as well as take advantage of
accelerations and optimizations.

Let us describe the class of functions to be optimized with some non-exclusive
properties:

\vspace*{-0.4em}
\begin{itemize} \itemsep -1pt
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  %\item {\bf numeric}: all elements of $x$ take values in $\mathcal{R}$
  \item {\bf sparse}: the gradient $f'(x)$ or $f_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the gradient $f'_j(x)$ is computable for
individual elements $x_j$ of $x$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}
\vspace*{-0.4em}

@@

In some cases, some subclasses of functions can still be optimized with more
general optimizers.  For example, separable functions can still be optimized
with an optimizer that does not specifically support them.  Similarly, a sparse
differentiable function may be optimized with any optimizer that supports
differentiable functions; however, the sparseness might not be taken advantage
of.

Now, with an understanding of the types of functions that we wish to support and
their individual characteristics, we can develop an API that a given function
can implement the relevant parts of.

\vspace*{-0.3em}
\section{{\tt FunctionType} API}
\vspace*{-0.2em}

In order to facilitate consistent implementations, we have defined a {\tt
FunctionType} API that describes all the methods that an objective function
may implement.  mlpack offers a few variations of this API, each designed to
cover some of the function characteristics of the previous section.
%At the time of writing, we offer a few implementations of this API, each
%designed to cover a special case.
%\begin{itemize}
%  \item{\tt FunctionType}: Requires the implementation of {\tt Evaluate()} and
%  {\tt Gradient()}. This is a generic function that is used by the
%  {\tt GradientDescent} optimizer.
%  \item{\tt DecomposableFunctionType}: In addition to {\tt FunctionType}'s
%  requirements, requires the implementation of {NumFunctions()}, which returns
%  the number of parts the optimization problem can be decomposed into. The other
%  two functions ({\tt Gradient(), Evaluate()}) accept an index as argument and
%  only evaluate one of the decomposable parts. This is used by our {\tt AdaGrad}
%  implementations, some {\tt SGD} implementations, {\tt RMSProp} and
%  {\tt SMORMS3}.
%  \item{\tt SparseFunctionType}: Similar to {\tt DecomposableFunctionType},
%  however this function type returns a sparse gradient, which makes it ideal for
%  sparse, parallelizable optimization problems. This function type is used by
%  {\bf mlpack}'s parallel implementation of SGD, {\tt ParallelSGD}.
% We'll kind of sweep constrained optimization under the rug a little bit, there
% are a lot of extra complexities we don't have space for.
%  \item{\tt LagrangianFunctionType}: This function type fits constrained
%  optimization problems. It requires the implementation of {\tt NumConstraints},
%  {\tt EvaluateConstraint()} and {\tt GradientConstraint}, which are required by
%  optimizers like {\tt AugLagrangian} and {\tt LRSDP}.
%\end{itemize}
Any {\tt FunctionType} to be optimized requires the implementation of an {\tt
Evaluate()} method.  The interface used for that can be one of the following
two methods:

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{c++}
// For non-separable objectives.
double Evaluate(const arma::mat& parameters);
// For separable objectives.
double Evaluate(const arma::mat& parameters,
                const size_t start,
                const size_t batchSize);
\end{minted}
\vspace*{-0.3em}

Both of these methods should calculate the objective given the parameters
matrix {\tt parameters}; however, the second overload is for {\it separable}
functions, and should calculate the partial objective starting at the separable
function indexed by {\tt start} and calculate {\tt batchSize} partial objectives
and return the sum.  Functions implementing the first overload are used by
optimizers like the {\tt GradientDescent} optimizer; functions implementing the
second are used by SGD-like optimizers.

Note that, importantly, it is easy to calculate the objective for a
non-separable function with the second overload just by setting {\tt start} to
$0$ and {\tt batchSize} to $1$.  Therefore, it is easy to make an `adapter' that
can allow separable optimizers to work with non-separable functions (and vice
versa).

Next, any {\it differentiable} function must implement some {\tt Gradient()}
method.

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{c++}
// For non-separable differentiable sparse and non-sparse functions.
template<typename GradType>
void Gradient(const arma::mat& parameters, GradType& gradient);
// For separable differentiable sparse and non-sparse functions.
template<typename GradType>
void Gradient(const arma::mat& parameters,
              const size_t start,
              GradType& gradient,
              const size_t batchSize);
\end{minted}
\vspace*{-0.3em}

Both of these methods should calculate the gradient and place the results into
the matrix object {\tt gradient}.  Note that the method accepts a template
parameter {\tt GradType}, which may be {\tt arma::mat} (dense Armadillo matrix)
or {\tt arma::sp\_mat} (sparse Armadillo matrix).  This allows support for both
sparse-supporting and non-sparse-supporting optimizers.\footnote{One could write
a non-templated {\tt Gradient()} method for just {\tt arma::mat}, and it would
work fine for non-sparse optimizers.  But to us it seems just as easy to
templatize it.}

Next, if the objective function is {\it partially differentiable}, we can
implement the following method:

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{c++}
// For partially differentiable sparse and non-sparse functions.
template<typename GradType>
void PartialGradient(const arma::mat& parameters,
                     const size_t j,
                     GradType& gradient);
\end{minted}
\vspace*{-0.3em}

This should calculate the gradient of the parameters {\tt parameters} with
respect to the parameter {\tt j} and store the results (either sparse or dense)
in the {\tt gradient} matrix object.

In addition, {\it separable} and {\it partially differentiable} functions must
implement the {\tt NumFunctions()} and {\tt NumFeatures()} functions,
respectively:

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{c++}
// For separable functions: return the number of parts the optimization problem
// can be decomposed into.
size_t NumFunctions();
// For partially differentiable functions: return the number of partial derivatives.
size_t NumFeatures();
\end{minted}
\vspace*{-0.3em}

Finally, {\it separable} functions must implement the method
{\footnotesize \mintinline{c++}{void Shuffle()}}, which shuffles the ordering
of the functions (note that the optimizer is not required to call it).  This is
useful for data-based problems, where it may not be desirable to loop over the
separable objective functions---which usually correspond to individual data
points---in the same order.

These simple functions, however, do not specify how to handle {\it categorical}
or {\it bounded} functions.  In those cases, the {\tt FunctionType} should
accept, in its constructor, its bounds and which dimensions are categorical or
numeric.  Since constraint types can differ greatly, it is up to an individual
optimizer to define the format in which it should receive its constraints.  More
information can be found in the documentation of mlpack's {\tt LRSDP}, {\tt
AugLagrangian}, and {\tt FrankWolfe} optimizers.

%% The constraints are part of the optimizer themselves.
% EvaluateConstraint()
% NumConstraints()

%\begin{itemize}
%  \item {\tt double Evaluate(const arma::mat& parameters)} {\it (arbitrary
%functions)}: return the objective function given the parameters.
%  \item {\tt void Gradient(const arma::mat& parameters, arma::mat& gradient)}
%{\it (differentiable)}: compute the gradient and store it in {\tt gradient}
%given the parameters.
%  \item {\tt double Evaluate(const arma::mat& parameters, const size\_t start,
%const size\_t batchSize)} {\it (separable)}: compute a partial objective
%function, starting with separable function {\tt start} and computing {\tt batchSize}
%functions.
%  \item {\tt void Gradient(arma::mat& parameters, const size\_t start,
%arma::mat& gradient, const size\_t batchSize)} {\it (separable)}: compute a part
%of the gradient starting with separable function {\tt start} and computing {\tt
%batchSize} individual gradients, storing the result in {\tt gradient}.

\vspace*{-0.3em}
\section{Optimizer API}
\vspace*{-0.4em}

In addition to implementing functions, users can also add new optimizers easily
if they implement an optimizer with a simple API.  Fortunately, the requirements
for implementing optimizers are much simpler than for objective functions.  An
optimizer must implement only the method

\vspace*{-0.5em}
\begin{minted}[fontsize=\footnotesize]{c++}
template<typename FunctionType>
double Optimize(FunctionType& function,
                arma::mat& parameters);
\end{minted}
\vspace*{-0.5em}

The {\tt Optimize()} method should check that the given {\tt FunctionType}
satisfies the assumptions the optimizer makes (see Section \ref{sec:static}) and
optimize the given function {\tt function}, storing the best set of parameters
in the matrix {\tt parameters} and returning the best objective value.

\vspace*{-0.3em}
\section{Example function and optimizer}
\vspace*{-0.4em}

@@

% \vspace*{-0.3em}
%\begin{listing}[H]
\begin{minted}[fontsize=\footnotesize]{c++}
class ObjectiveFunction {
  ObjectiveFunction() {
    in = arma::vec("20 12 15 100");
    bi = arma::vec("-4 -2 -3 -8");
  }

  size_t NumFunctions() { return 4; }
  void Shuffle() { order = arma::shuffle(arma::uvec("0 1 2 3 ")); }

  double Evaluate(const arma::mat& para, size_t s, size_t bs) {
    double cost = 0;
    for (size_t i = s; i < s + bs; i++)
      cost += (para(i) * para(i)) + bi(order(i)) * para(order(i)) + in(order(i));
    return cost;
  }

  void Gradient(const arma::mat& para, size_t s, size_t bs, arma::mat& g) {
    g.zeros(para.n_rows, para.n_cols);
    for (size_t i = s; i < s + bs; i++)
      g(i) += (1.0 / bs) * 2 * para(i) + bi(order(i));
  }

  arma::vec in, bi; arma::uvec order;
};
\end{minted}
%\vspace{-0.6em}
%\caption{Simple objective function where each dimension has a parabola with a
%  distinct minimum}
%\vspace{-0.8em}
%\end{listing}
%\vspace{-0.8em}
%\begin{listing}[H]
\begin{minted}[fontsize=\footnotesize]{c++}
class SimpleOptimizer {
  SimpleOptimizer(size_t bs = 2, double lr = 0.1) : bs(bs), lr(lr) { }

  template<typename FunctionType>
  double Optimize(FunctionType& function, arma::mat& parameter) {
    arma::mat gradient;
    for (size_t i = 0; i < 300; i += bs) {
      const size_t funIdx = i % 4;
      function.Gradient(parameter, funIdx, gradient, batchSize);
      parameter -= lr * gradient;
      if (funIdx == 0) { function.Shuffle(); }
    }
    return function.Evaluate(parameter, 0, function.NumFunctions());
  }
  size_t bs; double lr; // Store batch size and learning rate internally.
};
\end{minted}
%\vspace{-0.6em}
%\caption{Simple stochastic gradient descent method with a static number of
%iterations to optimize the function parameter ({\tt bs} defines the batch size
%and {\tt lr} the learning rate). )}
%\vspace{-0.6em}
%\end{listing}
%\vspace{-0.8em}
Using this code, we can compile the code, and when we run it

\vspace*{-0.4em}
\begin{minted}[fontsize=\footnotesize]{c++}
  ObjectiveFunction function; arma::mat parameter("0 0 0 0;")
  SimpleOptimizer optimizer;
  std::cout << optimizer.Optimize(function, parameter);
\end{minted}
\vspace*{-0.4em}

we receive the following output:

\vspace*{-0.4em}
{\footnotesize
\begin{verbatim}
xxxx
\end{verbatim}
}
\vspace*{-0.4em}

@@

\section{Statically checking function properties}
\label{sec:static}

Unfortunately, template metaprogramming can result in some very lengthy error
messages.  Therefore, we must be careful to ensure that a user is able to easily
debug a problem when they implement an objective function or gradient.  To
provide good error message output, we can use C++'s template metaprogramming
support to determine what methods a type has available.  Similarly, we can also
use static compile-time constants to denote the methods that are required by a
specific optimizer.  This is implemented via SFINAE \cite{sfinae}.  A {\tt
static\_assert()} is raised when a given objective function does not have the
required methods for a given optimizer.

For instance, when attempting to use the L-BFGS optimizer without having a
{\tt Gradient()} function implemented, the user will receive a (comparatively)
simple error message of the form:

\vspace*{-0.4em}
{\footnotesize
\begin{verbatim}
  error: static assertion failed: the FunctionType does not have a correct
      definition of a Gradient() function
\end{verbatim}
}
\vspace*{-0.4em}

@@

\vspace*{-0.4em}
% I guess to save some space we should group them.
\begin{itemize} \itemsep -1pt
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

  \item {\bf Other:} Conditional Gradient Descent, Frank-Wolfe algorithm

  \item {\bf Objective functions:} Neural Networks, Logistic regression,
      Matrix completion, Neighborhood Components Analysis, Regularized SVD,
      Reinforcement learning, Softmax regression, Sparse autoencoders,
      Sparse SVM
\end{itemize}
\vspace*{-0.4em}

@@

We have identified that the support for generic and robust optimization is not
currently available in most machine learning toolkits, and acted upon this
observation to provide an easy framework for both implementing new optimizers
and new objective functions to be optimized inside of the mlpack machine
learning library.

The framework provided by mlpack supports a wide array of special cases, and
already has implemented specialized algorithms that outperform their classic
generic alternatives.
